---
output:
  word_document:
    reference_docx: word-styles-reference-02.docx
    fig_caption: yes
---

```{r setup, include=FALSE}
### FILL IN THIS METADATA BEFORE RUNNING SCRIPT ###
reg = "PRIAs" # unabbreviated region
subreg_mar = "no"
## subreg_mar = "NMAR" or "SMAR" --- optional
survyr = "2018" # format: "20XX" (enter manually)
###################################################

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "C:/Users/Courtney.S.Couch/Documents/Courtney's Files/R Files/ESD/ASRAMP 2018")

# load library functions
source("C:/Users/Courtney.S.Couch/Documents/GitHub/Benthic-Scripts/Functions/mw_functions.R")
library(ggplot2)
library(plyr)
library(reshape2)
library(RColorBrewer)
library(vegan)
library(car)
library(dplyr)
library(gridExtra)
library(foreign)
library(haven)
library(knitr)
# rounds the y-axis labels
fmt <- function(){
  f <- function(x) as.character(round(x,2))
  f
}

# for figures:
element_grob.element_custom <- function(element, label="", ...)  {
  disect <- strsplit(label, "\\n")[[1]]
  g1 <- tableGrob(as.matrix(disect), parse=TRUE,
                  theme=theme.list(gpar.corefill = gpar(fill = NA, col = NA),
                                   core.just = "center"))
  gTree(children=gList(g1), height=grobHeight(g1), cl = "custom_strip")
}
# gTrees don't know their size and ggplot would condense it, so give it room
grobHeight.custom_strip = heightDetails.custom_axis = function(x, ...)
  x$height
# wrapper to fool ggplot2's checks...
facet_custom <- function(...){
  structure(
    list(...), # this ... information is not used
    class = c("element_custom","element_blank", "element") # inheritance test workaround
  ) 
}

# read in data
getwd()
surveysites <- read.csv(paste("C:/Users/Courtney.S.Couch/Documents/Courtney's Files/R Files/ESD/ASRAMP 2018/", reg, "_", survyr, "ASRAMP2018_Sitelist.csv", sep = ""))
area <- read.csv(paste("T:/Benthic/Data/REA Coral Demography/WorkingData_forBSRs/", reg, "_sectorArea.csv", sep = ""))
totscl_depth <- read.csv(paste("T:/Benthic/Data/REA Coral Demography/WorkingData_forBSRs/", reg, "_", survyr, "_IslandDepth_metrics_totalscl_frf.csv", sep = ""))
gen_isl <- read.csv(paste("T:/Benthic/Data/REA Coral Demography/WorkingData_forBSRs/", reg, "_", survyr, "_Islandmetrics_targetgenera_frf.csv", sep = ""))
totscl_isl <- read.csv(paste("T:/Benthic/Data/REA Coral Demography/WorkingData_forBSRs/", reg, "_", survyr, "_Islandmetrics_totalscl_frf.csv", sep = ""))
allgen_isl <- rbind(gen_isl, totscl_isl)
      # island order: (for figures)
      islaord <- island_order(reg)
  allgen_isl <- allgen_isl[ order(match(allgen_isl$ISLAND, islaord)),]
  allgen_isl$ISLAND<-as.character(allgen_isl$ISLAND)
  allgen_isl$ISLAND<-factor(allgen_isl$ISLAND, levels = islaord)
# sp_isl <- read.csv(paste("T:/Benthic/Data/REA Coral Demography/WorkingData_forBSRs/", reg, "_", survyr, "_Islandmetrics_targettaxon_frf.csv", sep = ""))
#   totscl_isl2 <- totscl_isl
#   colnames(totscl_isl2)[5] <- "SPCODE"  
#   allsp_isl <- rbind(sp_isl, totscl_isl2)
#   allsp_isl <- allsp_isl[ order(match(allsp_isl$ISLAND, islaord)),]
#   allsp_isl$ISLAND<-as.character(allsp_isl$ISLAND)
#   allsp_isl$ISLAND<-factor(allsp_isl$ISLAND, levels = islaord)
#   head(allsp_isl)

## variables to fill
# survey dates: (survdates)
surveysites$DATE_ <- as.Date(surveysites$DATE_)
  mind <- min(surveysites$DATE_)
  maxd <- max(surveysites$DATE_)
    survdates = paste(format(mind, "%B"), format(mind, "%d"), "-", format(maxd, "%B"), format(maxd, "%d"), sep = " ") # survey dates- format: "Month Day - Month Day"
# genus full names and title: (for figures)
genera <- unique(gen_isl[ which(gen_isl$GENUS_CODE != "SSSS"),]$GENUS_CODE)
gen1 <- gen_name_fun(genera[1])
gen2 <- gen_name_fun(genera[2])
gen3 <- gen_name_fun(genera[3])
gen4 <- gen_name_fun(genera[4])
title_gen <- c(substitute(expr = italic(gen1), env = list(gen1=gen1)), substitute(expr = italic(gen2), env = list(gen2=gen2)), substitute(expr = italic(gen3), env = list(gen3=gen3)), substitute(expr = italic(gen4), env = list(gen4=gen4)), "plain('       Total\nScleractinians')")
# species full names and title: (for figures)
species <- unique(sp_isl[ which(sp_isl$SPCODE != "SSSS"),]$SPCODE)
spe1 <- spe_name_fun(species[1])
spe2 <- spe_name_fun(species[2])
spe3 <- spe_name_fun(species[3])
spe4 <- spe_name_fun(species[4])
title_sp <- c(substitute(expr = italic(spe1), env = list(spe1=spe1)), substitute(expr = italic(spe2), env = list(spe2=spe2)), substitute(expr = italic(spe3), env = list(spe3=spe3)), substitute(expr = italic(spe4), env = list(spe4=spe4)), "plain('       Total\nScleractinians')")

```
# Benthic monitoring summary: `r reg_name(reg)` `r survyr`

## About this summary
The purpose of this document is to provide a brief summary of the most recent survey efforts performed in the `r reg_name(reg)` in `r survyr` by the Ecosystem Sciences Division (ESD) of the NOAA Pacific Islands Fisheries Science Center as part of the Pacific Reef Assessment and Monitoring Program (Pacific RAMP). A more detailed assessment of the coral populations and reef community structure from the `r reg_name(reg)` will be summarized in future publications.

## Sampling effort 
* Ecological monitoring in the `r reg_name(reg)` was performed from `r survdates` `r survyr`.
* Surveys were conducted at `r nrow(surveysites)` sites across `r length(unique(surveysites$ISLAND))` islands: `r unique(surveysites$ISLAND)`.
* Coral demography, partial mortality, and condition were surveyed using belt transects; benthic community structure will be assessed from photoquadrat digital images.

###### Table 1. Reef zone area (hectare) and number of sites surveyed (n) for the three depth cateogires surveyed in the `r reg_name(reg)`.
```{r, echo = FALSE}
# produce a table showing the number of sites by depth strata, as well as the area (km2) of each strata (report all reef zones here but only forereef data is shown in the figures)

# area table
area$REEF_ZONE <- as.character(area$REEF_ZONE)
area$REEF_ZONE[area$REEF_ZONE == "PRS"] <- "Forereef"
area$REEF_ZONE <- as.factor(area$REEF_ZONE)
area_ha <- aggregate(AREA_HA ~ ISLAND + REEF_ZONE + DEPTH_BIN, data = area, FUN = sum)
area_ha <- spread(area_ha[,c("REEF_ZONE", "ISLAND", "DEPTH_BIN", "AREA_HA")], DEPTH_BIN, AREA_HA)
colnames(area_ha) <- c("Zone", "Island", "Deep", "Mid", "Shallow")
area_ha[, 3:5][is.na(area_ha[, 3:5])] <- 0
area_ha <- area_ha[ order(match(area_ha$Zone, zone_ord)),]

# site table
surveysites$REEF_ZONE <- as.character(surveysites$REEF_ZONE)
surveysites$REEF_ZONE[surveysites$REEF_ZONE == "PRS" | surveysites$REEF_ZONE == "Protected Slope"] <- "Forereef"
surveysites$REEF_ZONE <- as.factor(surveysites$REEF_ZONE)
surveysites <- aggregate(SITE ~ ISLAND + REEF_ZONE + DEPTH_BIN, data = surveysites, FUN = length)
site_n <- spread(surveysites[,c("REEF_ZONE", "ISLAND", "DEPTH_BIN", "SITE")], DEPTH_BIN, SITE)
colnames(site_n) <- c("Zone", "Island", "ndeep", "nmid", "nshallow")
site_n[, 3:5][is.na(site_n[, 3:5])] <- 0
site_n <- site_n[ order(match(site_n$Zone, zone_ord)),]

# merge the area and site tables together
area_n <- merge(area_ha, site_n, by = c("Island", "Zone"))
area_n <- area_n[c(2,1,5,8,4,7,3,6)]
area_n <- area_n[ order(match(area_n$Zone, zone_ord)),]

print(area_n)
```

## Overview of data collected 

```{r, echo = FALSE}
### fig. 1, 2: bubble plot coral dens by site
# produce bubble plots of adult and juvenile coral density at survey sites in ArcMap - will use site level metrics
```

###### Figure 1. Mean density of all adult coral colonies (Scleractinia, >= 5 cm) at survey sites.  

###### Figure 2. Mean density of all juvenile coral colonies (Scleractinia, < 5 cm) at survey sites.  

```{r, warning = FALSE, message = FALSE, echo = FALSE, fig.width=3.78, fig.height=3.13}
### fig. 3: bar graph coral dens by strata

totscl_depth <- totscl_depth[,c("ISLAND", "DEPTH_BIN", "meanAdultColDen", "meanJuvColDen", "seAdultColDen", "seJuvColDen"),] # drop unnecessary columns

totscl_depth <- melt(setDT(totscl_depth), 
     measure = patterns("mean", "se"),
     variable.name = 'var', value.name = c('avg', 'se'))
totscl_depth$var <- ifelse( totscl_depth$var == 1, "Adult", "Juv")

totscl_depth <- totscl_depth[ order(match(totscl_depth$ISLAND, islaord)),]
totscl_depth$ISLAND<-as.character(totscl_depth$ISLAND)
totscl_depth$ISLAND<-factor(totscl_depth$ISLAND, levels = islaord)
levels(totscl_depth$DEPTH_BIN) <- c("Shallow", "Mid", "Deep")

Fig3plot <- ggplot(totscl_depth, aes(x=ISLAND, y=avg, fill=var)) + 
  geom_bar(position=position_dodge(), stat="identity", color="black") + 
  facet_grid(DEPTH_BIN ~ .) + 
  theme_bw()  + 
  theme(plot.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.ticks.x = element_blank(),
        axis.title.x = element_text(vjust = -.0001)) +
  scale_x_discrete(expand = c(0.005, 0.005)) + 
  scale_y_continuous(labels = fmt()) + 
  theme(text = element_text(size=10)) + 
  theme(strip.text.y = element_text(angle=0)) + 
  geom_errorbar(aes(ymin=avg-se, ymax=avg+se),width=.15, position=position_dodge(.9)) + 
  ylab(expression(bold(Density~~"(no. colonies"*~m^"-2"*")"))) + xlab(expression(bold(Island))) + scale_fill_manual(values=c("darkblue", "turquoise")) + guides(fill=guide_legend(title=NULL)) + theme(legend.position="top")

if(subreg_mar == "NMAR"){
  Fig3plot <- Fig3plot +
      theme(axis.text.x = element_text(angle = 40, hjust = 1))
}

Fig3plot
```

###### Figure 3. Mean density of adult (dark) and juvenile (light) coral colonies (± standard error) grouped by island within three depth categories: (A) shallow (0-6 m), (B) mid (>6-18 m), and (C) deep (>18-30).

```{r, echo = FALSE, fig.width=3.78, fig.height=3.13}
### fig. 4: bar graph coral dens by genera (adults)
allgen_isl_ads <- allgen_isl[,c("ISLAND", "GENUS_CODE", "meanAdultColDen", "seAdultColDen"),] # drop unnecessary columns
allgen_isl_ads$GENUS_CODE <- factor(allgen_isl_ads$GENUS_CODE, labels=title_gen)

Fig4plot <- ggplot(allgen_isl_ads, aes(x=ISLAND, y=meanAdultColDen, fill=GENUS_CODE)) + geom_bar(position=position_dodge(), stat="identity", color="black") + 
  guides(fill=FALSE) + facet_grid(GENUS_CODE ~ ., scales="free", labeller=label_parsed) + theme_bw() +
  theme(
    plot.background = element_blank()
    ,panel.grid.major = element_blank()
    ,panel.grid.minor = element_blank()
    ,axis.ticks.x = element_blank() # no x axis ticks
    ,axis.title.x = element_text( vjust = -.0001) # adjust x axis to lower the same amount as the genus labels
    ,legend.position="none"
  ) + scale_x_discrete(expand = c(0.005, 0.005)) + scale_y_continuous(labels = fmt()) + theme(text = element_text(size=10)) + theme(strip.text.y = element_text(angle=0)) + 
  geom_errorbar(aes(ymin=meanAdultColDen-seAdultColDen, ymax=meanAdultColDen+seAdultColDen),width=.15, position=position_dodge(.9)) + 
  ylab(expression(bold(Density~~"(no. colonies"*~m^"-2"*")"))) + xlab(expression(bold(Island))) + 
  scale_fill_manual(values=c("red3", "wheat3", "aquamarine3", "turquoise4", "darkblue")) 

if(subreg_mar == "NMAR"){
  Fig4plot <- Fig4plot +
      theme(axis.text.x = element_text(angle = 40, hjust = 1))
}

Fig4plot
```

###### Figure 4. Mean island-wide density of adult colonies (± SE) for total scleractinians and the four most abundant genera: `r strsplit(gen1," ")[[1]][1]`, `r strsplit(gen2," ")[[1]][1]`, `r strsplit(gen3," ")[[1]][1]`, and `r strsplit(gen4," ")[[1]][1]`.

```{r, echo = FALSE, fig.width=3.78, fig.height=3.13}
### fig. 5: bar graph coral dens by genera (juveniles)
allgen_isl_juv <- allgen_isl[,c("ISLAND", "GENUS_CODE", "meanJuvColDen", "seJuvColDen"),] # drop unnecessary columns
allgen_isl_juv$GENUS_CODE <- factor(allgen_isl_juv$GENUS_CODE, labels=title_gen)

Fig5plot <- ggplot(allgen_isl_juv, aes(x=ISLAND, y=meanJuvColDen, fill=GENUS_CODE)) + geom_bar(position=position_dodge(), stat="identity", color="black") + 
  guides(fill=FALSE) + facet_grid(GENUS_CODE ~ ., scales="free", labeller=label_parsed) + theme_bw() +
  theme(
    plot.background = element_blank()
    ,panel.grid.major = element_blank()
    ,panel.grid.minor = element_blank()
    ,axis.ticks.x = element_blank() # no x axis ticks
    ,axis.title.x = element_text( vjust = -.0001) # adjust x axis to lower the same amount as the genus labels
    ,legend.position="none"
  ) + scale_x_discrete(expand = c(0.005, 0.005)) + scale_y_continuous(labels = fmt()) + theme(text = element_text(size=10)) + theme(strip.text.y = element_text(angle=0)) + 
  geom_errorbar(aes(ymin=meanJuvColDen-seJuvColDen, ymax=meanJuvColDen+seJuvColDen),width=.15, position=position_dodge(.9)) + 
  ylab(expression(bold(Density~~"(no. colonies"*~m^"-2"*")"))) + xlab(expression(bold(Island))) + 
  scale_fill_manual(values=c("red2", "wheat2", "aquamarine2", "turquoise3", "blue3"))  

if(subreg_mar == "NMAR"){
  Fig5plot <- Fig5plot +
      theme(axis.text.x = element_text(angle = 40, hjust = 1))
}

Fig5plot
```

###### Figure 5. Mean density of juvenile coral colonies (± standard error) across islands grouped by total scleractinians and the four most abundant genera: `r strsplit(gen1," ")[[1]][1]`, `r strsplit(gen2," ")[[1]][1]`, `r strsplit(gen3," ")[[1]][1]`, and `r strsplit(gen4," ")[[1]][1]`."  

```{r, echo = FALSE, fig.width=3.78, fig.height=3.13}
### fig. 6: bar graph coral dens by species (adults)
allsp_isl_ads <- allsp_isl[,c("ISLAND", "SPCODE", "meanAdultColDen", "seAdultColDen"),] # drop unnecessary columns
head(allsp_isl_ads)
allsp_isl_ads$SPCODE <- factor(allsp_isl_ads$SPCODE, labels=title_sp)

Fig6plot <- ggplot(allsp_isl_ads, aes(x=ISLAND, y=meanAdultColDen, fill=SPCODE)) + geom_bar(position=position_dodge(), stat="identity", color="black") + 
  guides(fill=FALSE) + facet_grid(SPCODE ~ ., scales="free", labeller=label_parsed) + theme_bw() +
  theme(
    plot.background = element_blank()
    ,panel.grid.major = element_blank()
    ,panel.grid.minor = element_blank()
    ,axis.ticks.x = element_blank() # no x axis ticks
    ,axis.title.x = element_text( vjust = -.0001) # adjust x axis to lower the same amount as the genus labels
    ,legend.position="none"
  ) + scale_x_discrete(expand = c(0.005, 0.005)) + scale_y_continuous(labels = fmt()) + theme(text = element_text(size=10)) + theme(strip.text.y = element_text(angle=0)) + 
  geom_errorbar(aes(ymin=meanAdultColDen-seAdultColDen, ymax=meanAdultColDen+seAdultColDen),width=.15, position=position_dodge(.9)) + 
  ylab(expression(bold(Density~~"(no. colonies"*~m^"-2"*")"))) + xlab(expression(bold(Island))) + 
  scale_fill_brewer(palette="Set1")

if(subreg_mar == "NMAR"){
  Fig6plot <- Fig6plot +
      theme(axis.text.x = element_text(angle = 40, hjust = 1))
}

Fig6plot
```

###### Figure 6. Mean island-wide density of adult coral colonies (± standard error) for the four most abundant species and total scleractinians.  

*Coral condition and threatened species*- The percent of adult coral colonies exhibiting signs of bleaching, disease, or crown-of-thorns sea star (COTS) predation lesions are shown in Table 1. 'Disease' was defined as the combination of all identified diseases and lesions. 'Bleaching' included all levels of bleaching extent and severity. 'COTS' was defined as recent coral mortality attributable to COTS predation. The Endangered Species Act (ESA) threatened species [*insert spp. name*] was observed at [*insert islands*].    
 
###### Table 2. Percent of adult coral colonies that exhibited bleaching, disease, and crown-of-thorns Sea star (COTS) predation lesions in the `r reg`.  

```{r echo=FALSE}
### bleaching, disease, COTS prevalence - tab. 2
tab2 <- totscl_isl[,c(4,8,9,12,15,16,19)]
tab2 <- tab2[c(1,2,5,3,6,4,7)]
tab2 <- cbind(tab2[1], round(tab2[-1], 3))

print(tab2)
```

## Survey sampling design  
A single stage stratified random sampling design was employed to survey the `r regname`. The stratification scheme incorporated all hard bottom habitats located on the fore reef, and sampled across three depth categories: shallow (0-6 m), mid (>6-18 m) and deep (>18-30 m).  Allocation of sampling effort was proportional to total strata area.  Sites (geographic coordinates) were randomly selected within each stratum.

## Survey methods  
Coral demography surveys at each site were conducted along one belt transect.  Adult coral colonies (>= 5 cm) were surveyed within 10 m2 on each transect. Colonies were identified to the lowest taxonomic level possible (species or genus), measured (maximum diameter to the nearest cm), and morphology was noted.  In addition, partial mortality and condition of each colony was assessed.  Partial colony mortality was quantified as the percent of dead tissue (classified as 'old dead' or 'recent dead'), and attributed to cause of mortality when known.  Conditions affecting each colony (i.e., disease and bleaching) were noted, along with the extent (percent of colony affected) and severity (ranging from moderate to acute). 

```{r fig.width=3.78, fig.height=2.1,echo=FALSE}
library(png)
library(grid)
img <- readPNG("T:/Benthic/Ecological Assessment/benthic_reports/summary_report/BSR_GUIDELINES/Fig7.png")
 grid.raster(img)
```

###### Figure 7.  Schematic of the method used to conduct demographic surveys of coral populations and benthic community structure at the Rapid Ecological Assessment (REA) sites.

Juvenile coral colonies (< 5 cm) were surveyed within 3 m2 on each transect. Juvenile colonies were identified in the field by a distinct tissue and skeletal boundary that distinguished them from fragments of larger adult colonies.  Each juvenile colony was identified to the lowest taxonomic level possible (genus or species) and measured (both the maximum and perpendicular diameter to the nearest 2 mm).  

The present summary focuses only on colony density and initial estimates of bleaching, disease, and COTS predation occurrence.  The island-scale estimates presented here are generated from site-level means and are not weighted by reef area within depth strata.  Our data also include island and strata estimates of coral size structure, frequency of occurrence, total colony abundance, mean proportion of partial mortality, and condition (e.g. disease and bleaching) prevalence. Estimates for strata are generated from site means. Forthcoming island-scale estimates (means and totals) will be weighted by area. These statistical survey design estimates of the full suite of coral population and community metrics will be included in future publications.  

## About the monitoring program  
Pacific RAMP forms a key part of the National Coral Reef Monitoring Program of NOAA's Coral Reef Conservation Program (CRCP), providing integrated, consistent, and comparable data across US Pacific islands and atolls. CRCP monitoring efforts have these aims:  

-	Document the status of reef species of ecological and economic importance
-	Track and assess  the status and trends of US coral reef ecosystems in response to environmental stressors and human activities
-	Evaluate the effectiveness of specific management strategies and identify actions for future and adaptive responses

In addition to the coral population and benthic community surveys outlined here, Pacific RAMP efforts include interdisciplinary monitoring of oceanographic conditions, fish population and assemblages, invertebrate diversity and abundance, coral reef habitat assessments and mapping, and studies of the effects of climate change and ocean acidification.  Data are available upon request.  

## For more information
Coral Reef Conservation Program:  
http://coralreef.noaa.gov  
NMFS Pacific Islands Fisheries Science Center:  
http://www.pifsc.noaa.gov  
Related publications:  
http://www.pifsc.noaa.gov/pubs/credpub.php  
Additional information:  
http://www.pifsc.noaa.gov/cred/benthic_monitoring.php  
Benthic team lead:  
bernardo.vargasangel@noaa.gov  
Benthic survey data requests:  
dione.swanson@noaa.gov


